---
title: "HDD_BPL"
author: "Barbara Pavlek Löbl"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Human designed dimensions

The goal of this analysis is to replicate the process of formation of "composite variables" in Seshat, by aggregation of 51 (mostly binary) variables into 9 aggregated classes ("complexity characteristics" ie. CCs: cf. <https://www.pnas.org/doi/full/10.1073/pnas.1708800115>):
  1. polity population (CC1)
  2. polity territory (CC2)
  3. capital population (size of the largest urban centre) (CC3)
  4. hierarchical complexity (CC4)
    -> administrative levels, military levels, settlement hierarchy,        religious levels
  5. government (CC5)
    -> officers, bureaurcrats, court, merit promotion, soldiers,          lawyers, judges, priests, exam system, legal code, government       buildings
  6. infrastructure (CC6)
    -> bridges, ports, roads, market, water supply, canals, mines,        irrigation, food storage
  7. information system (CC7)
    -> mnemonic, lists, records, script, alphabet, non-phonetic
  8. texts/writing (CC8)
    -> calendar, sacred texts, religious texts, practical texts,        science texts, history, fiction, philosophy
  9. monetary system (CC9)
    -> articles, tokens, metals, foregin coins, indigenous coins,       paper currency

In the article, it was hypothesized that the social complexity will be captured by two PCs, where the first (PC1) was predicted to be a combination of the first 4 CCs (CC1-4) - a metrics of "scale", and the second (PC2) would represent "non-scale" components (CC5-9).

Here we call these aggregated classes "human designed dimensions" (HDDs), as these categories have been created through the iterative coding process overseen by domain experts.

Based on the hierarchical representation of variables in the Seshat Cdebook, 51 variables are aggregated into HDDs over polities at each timestamp, taking the average value across the variables.


## Setting up

The starting dataset is based on the original Seshat data (Equinox-2020, <https://seshatdatabank.info/datasets/>), processed by Petr Tureček. All continious variables are converted to binary, and missing data is filled with arithmetic mean of the values of the whole variable.

```{r}
data_numbers_filled <- read.delim2("data_numbers_filled.txt", header = T, sep = "\t",
          stringsAsFactors=F,encoding="UTF-8")
d_Equinox <- data.frame(data_numbers_filled)
d_Equinox[5:128] <- lapply(d_Equinox[5:128], as.numeric)
```

For the list of variables, I relied on the Seshat Codebook (<https://seshatdatabank.info/wp-content/uploads/2020/04/April-15-Codebook-PDF.pdf>),and the list of "canonical variables" provided as a supplementary material of the cited PNAS paper:

```{r}
CanonVars <- read.delim2("CanonicalVars.txt", header = T, sep = "\t",
          stringsAsFactors=F,encoding="UTF-8")
d_Canon <- data.frame(CanonVars)
```

There are some discrepancies in the number of variables provided by the Equinox dataset (124), the Seshat Codebook (250+), and the variables list (164) and the Canonical Variables list (121) from the SI of the PNAS paper. Because most of the variables in the Canon correspond well with the ones in the processed Equinox dataset, the Canon variables and categories were used as a reference for aggregation.

## Aggregating variables into dimensions - original 9 CCs

Replicating original 9 "complexity characteristics" (CCs) from the PNAS paper (see above).

# 1.1 Exploring component variables - coverage
```{r}
which(colSums(d_Equinox[5:128]) ==0)
```

#1.2 Exploring component variables - correlation matrices
```{r}

#HComplex_CC4
HComplex_CC4 <- d_Equinox[,9:12]
print(cor(HComplex_CC4))

#Gov_CC5
Gov_CC5 <- cbind.data.frame(d_Equinox$Fulltime.bureaucrats,
                 d_Equinox$Professional.military.officers,
                 d_Equinox$Courts,
                 d_Equinox$Merit.promotion,
                 d_Equinox$Professional.soldiers,
                 d_Equinox$Formal.legal.code,
                 d_Equinox$Specialized.government.buildings,
                 d_Equinox$Professional.Lawyers,
                 d_Equinox$Judges,
                 d_Equinox$Professional.priesthood,
                 d_Equinox$Examination.system)

print(cor(Gov_CC5))

#Infrastr_CC6
Infrastr_CC6 <- cbind.data.frame(d_Equinox$Bridges,
                 d_Equinox$Ports,
                 d_Equinox$Roads,
                 d_Equinox$markets,
                 d_Equinox$drinking.water.supply.systems,
                 d_Equinox$Canals,
                 d_Equinox$Mines.or.quarries,
                 d_Equinox$irrigation.systems,
                 d_Equinox$food.storage.sites)

print(cor(Infrastr_CC6))

#InfoSys_CC7
InfoSys_CC7 <- cbind.data.frame(d_Equinox$Mnemonic.devices,
                 d_Equinox$Lists.tables.and.classifications,
                 d_Equinox$Written.records,
                 d_Equinox$Nonwritten.records,
                 d_Equinox$Script,
                 d_Equinox$Phonetic.alphabetic.writing,
                 d_Equinox$Nonphonetic.writing)

print(cor(InfoSys_CC7))

#Texts_CC8
Texts_CC8 <- cbind.data.frame(d_Equinox$Calendar,
                 d_Equinox$Sacred.Texts,
                 d_Equinox$Religious.literature,
                 d_Equinox$Practical.literature,
                 d_Equinox$Scientific.literature,
                 d_Equinox$History,
                 d_Equinox$Fiction,
                 d_Equinox$Philosophy)

print(cor(Texts_CC8))

#Money_CC9
Money_CC9 <- cbind.data.frame(d_Equinox$Articles,
                 d_Equinox$Tokens,
                 d_Equinox$Precious.metals,
                 d_Equinox$Foreign.coins,
                 d_Equinox$Indigenous.coins,
                 d_Equinox$Paper.currency)

print(cor(Money_CC9))

```

#1.3 Aggregating component variables - average
```{r}
HComplex_CC4_agg <- apply(HComplex_CC4, 1, mean)
Gov_CC5_agg <- apply(Gov_CC5, 1, mean)
Infrastr_CC6_agg <- apply(Infrastr_CC6, 1, mean)
InfoSys_CC7_agg <- apply(InfoSys_CC7, 1, mean)
Texts_CC8_agg <- apply(Texts_CC8, 1, mean)
Money_CC9_agg <- apply(Money_CC9, 1, mean)
```

#1.4. Full HDD dataset (9 original CCs)
```{r}
fullHDD_9CCs <- cbind.data.frame(d_Equinox[1:7],HComplex_CC4_agg,Gov_CC5_agg,Infrastr_CC6_agg,InfoSys_CC7_agg,Texts_CC8_agg,Money_CC9_agg)
colnames(fullHDD_9CCs)[5:7] <- c("PolPop_CC1","PolTerr_CC2","CapPol_CC3")

write.table(fullHDD_9CCs,"HDD_9CCs_BPL.txt",sep="\t",row.names=F)
```

## Plotting 

Dimensions values for each of the 414 polities in 30 NGAs over time, 100-y time intervals.

*TBD*

```{r ggplot, echo=FALSE}

```


## Aggregating variables into dimensions - Canonical Variables Categories & Subsections

Creating different HDDs based on the hierarchical structure of variables in the Seshat as presented in the Canonical Variables dataset (columns "Subsection" and "Category"). 

Canon-Subsections

```{r}
unique(d_Canon[,4])
```

[1] "Social Scale" => Scale        
[2] "Hierarchical Complexity" => HComplex

[3] "Professions" => Gov                                          
[4] "Bureaucracy characteristics" => Gov                         
[11] "Status" ("elite status is hereditary") => Gov
[14] "Limits on Power of the Chief Executive"  => Gov

[5] "Law" => Law 

[6] "Specialized Buildings: polity owned" => Infrastr

[7] "Other" (Category: money) => Money                                                
[7] "Other" (Category: infrastructure / postal services) => Info
[8] "Information" (Category: texts & writing) => Info

[9] "MilTech" => MilTech                                         
[10] "Fortifications" => MilTech 

[12] "Normative Ideological Aspects of Equity and Prosociality" => Norms
[13] "Deification of Rulers" => Norms
               
[15] "Moralizing Supernatural Powers"  => MoralGods


#1.1 Exploring component variables - correlation matrices
```{r}
Scale <- cbind.data.frame(d_Equinox$Polity.Population,
                 d_Equinox$Polity.territory,
                 d_Equinox$Population.of.the.largest.settlement)
print(cor(Scale))

HComplex <- cbind.data.frame(d_Equinox$Administrative.levels,
                 d_Equinox$Military.levels,
                 d_Equinox$Religious.levels,
                 d_Equinox$Settlement.hierarchy)
print(cor(HComplex))

Gov <- cbind.data.frame(d_Equinox$Professional.military.officers,
                 d_Equinox$Professional.soldiers,
                 d_Equinox$Professional.priesthood,
                 d_Equinox$Fulltime.bureaucrats,
                 d_Equinox$Examination.system,
                 d_Equinox$Merit.promotion,
                 d_Equinox$Specialized.government.buildings)
print(cor(Gov))

Law <- cbind.data.frame(d_Equinox$Courts,
                 d_Equinox$Formal.legal.code,
                 d_Equinox$Judges,
                 d_Equinox$Professional.Lawyers)
print(cor(Law))

Infrastr <- cbind.data.frame(d_Equinox$irrigation.systems,
                 d_Equinox$drinking.water.supply.systems,
                 d_Equinox$markets,
                 d_Equinox$food.storage.sites,
                 d_Equinox$Roads,
                 d_Equinox$Bridges,
                 d_Equinox$Canals,
                 d_Equinox$Ports,
                 d_Equinox$Mines.or.quarries)
print(cor(Infrastr))

Money <- cbind.data.frame(d_Equinox$Articles,
                 d_Equinox$Tokens,
                 d_Equinox$Precious.metals,
                 d_Equinox$Foreign.coins,
                 d_Equinox$Indigenous.coins,
                 d_Equinox$Paper.currency)
print(cor(Money))


Info <- cbind.data.frame(d_Equinox$Couriers,
                 d_Equinox$Postal.stations,
                 d_Equinox$General.postal.service,
                 d_Equinox$Mnemonic.devices,
                 d_Equinox$Nonwritten.records,
                 d_Equinox$Written.records,
                 d_Equinox$Nonphonetic.writing,
                 d_Equinox$Phonetic.alphabetic.writing,
                 d_Equinox$Lists.tables.and.classifications,
                 d_Equinox$Calendar,
                 d_Equinox$Sacred.Texts,
                 d_Equinox$Religious.literature,
                 d_Equinox$Scientific.literature,
                 d_Equinox$Practical.literature,
                 d_Equinox$History,
                 d_Equinox$Philosophy,
                 d_Equinox$Fiction)
print(cor(Info))

MilTech <- cbind.data.frame(d_Equinox$Copper,
                 d_Equinox$Bronze,
                 d_Equinox$Iron,
                 d_Equinox$Steel,
                 d_Equinox$Javelins,
                 d_Equinox$Atlatl,
                 d_Equinox$Slings,
                 d_Equinox$Self.bow,
                 d_Equinox$Composite.bow,
                 d_Equinox$Crossbow,
                 d_Equinox$Tension.siege.engines,
                 d_Equinox$Gunpowder.siege.artillery,
                 d_Equinox$Handheld.firearms,
                 d_Equinox$War.clubs,
                 d_Equinox$Battle.axes,
                 d_Equinox$Daggers,
                 d_Equinox$Swords,
                 d_Equinox$Spears,
                 d_Equinox$Polearms,
                 d_Equinox$Dogs,
                 d_Equinox$Donkeys,
                 d_Equinox$Horses,
                 d_Equinox$Elephants,
                 d_Equinox$Camels,
                 d_Equinox$Wood.bark.etc,
                 d_Equinox$Leather.cloth,
                 d_Equinox$Shields,
                 d_Equinox$Helmets,
                 d_Equinox$Breastplates,
                 d_Equinox$Limb.protection,
                 d_Equinox$Chainmail,
                 d_Equinox$Scaled.armor,
                 d_Equinox$Laminar.armor,
                 d_Equinox$Small.vessels.canoes.etc,
                 d_Equinox$Merchant.ships.pressed.into.service,
                 d_Equinox$Specialized.military.vessels,
                 d_Equinox$Settlements.in.a.defensive.position,
                 d_Equinox$Wooden.palisades,
                 d_Equinox$Earth.ramparts,
                 d_Equinox$Ditch,
                 d_Equinox$Moat,
                 d_Equinox$Stone.walls.mortared,
                 d_Equinox$Stone.walls.mortared,
                 d_Equinox$Fortified.camps,
                 d_Equinox$Complex.fortifications,
                 d_Equinox$Long.walls,
                 d_Equinox$Modern.fortifications)
print(cor(MilTech))

Norms <- cbind.data.frame(d_Equinox$Ideological.reinforcement.of.equality,
                 d_Equinox$Ideology.reinforces.prosociality,
                 d_Equinox$Ideological.thought.equates.elites.and.commoners,
                 d_Equinox$Ideological.thought.equates.rulers.and.commoners,
                 d_Equinox$production.of.public.goods)
print(cor(Norms))

MoralGods <- cbind.data.frame(d_Equinox$Moralizing.religion.adopted.by.elites,
                 d_Equinox$Moralizing.enforcement.in.this.life,
                 d_Equinox$Moralizing.enforcement.is.agentic,
                 d_Equinox$Moralizing.enforcement.of.rulers,
                 d_Equinox$Moralizing.enforcement.is.certain,
                 d_Equinox$Moral.concern.is.primary,
                 d_Equinox$Moralizing.enforcement.is.targeted,
                 d_Equinox$Moralizing.norms.are.broad,
                 d_Equinox$Moralizing.enforcement.in.afterlife,
                 d_Equinox$Moralizing.religion.adopted.by.commoners)
print(cor(MoralGods))


```

#1.2 Aggregating component variables - average
```{r}
Scale_agg <- apply(Scale, 1, mean)
HComplex_agg <- apply(HComplex, 1, mean)
Gov_agg <- apply(Gov, 1, mean)
Law_agg <- apply(Law, 1, mean)
Infrastr_agg <- apply(Infrastr, 1, mean)
Money_agg <- apply(Money, 1, mean)
Info_agg <- apply(Info, 1, mean)
MilTech_agg <- apply(MilTech, 1, mean)
Norms_agg <- apply(Norms, 1, mean)
MoralGods_agg <- apply(MoralGods, 1, mean)
```

#1.3. Full HDD dataset (10 categories)
```{r}
fullHDD_10cat <- cbind.data.frame(d_Equinox[1:4],Scale_agg,HComplex_agg,Gov_agg,Law_agg,Infrastr_agg,Money_agg,Info_agg,MilTech_agg,Norms_agg,MoralGods_agg)

write.table(fullHDD_10cat,"HDD_10cat_BPL.txt",sep="\t",row.names=F)
```

## Plotting 

Dimensions values for each of the 414 polities in 30 NGAs over time, 100-y time intervals.

*TBD*

```{r pressure, echo=FALSE}
```